Phonetic Transcription by Untrained Annotators

### 2018-06-03 2018-itat-krompachy/paper.tex
wip: itat 2018 text

\institute{ÚFAL MFF UK,\\
\email{kruza@ufal.mff.cuni.cz},\\ WWW home page:
\texttt{http://sixtease.net/}}

@@

\subsection{Representation}

@@

The module includes two-way transduction, although only the one from array of
phonemes to human-readable phonetic respelling is needed in our application.
Note that when encoding into the phonetic respelling, none of
{\em di ti ni dě tě ně} is ever output. The palatal consonants are always
explicitly spelled out.

The ambiguity of diphtong / subsequent vocals, as in {\em poučit [p o u ch i t]}
vs. {\em louže [l ow zh e]} is explicated by an apostrophe in the respelling
{\em po'učit / louže}. Voiceless alveolar fricative trill is explicated as {\em
r'} as in {\em tř'ít} vs. {\em dřít}.

>>>>

\institute{ÚFAL MFF UK,\\
\email{kruza@ufal.mff.cuni.cz}}

@@

\subsection{Phonetic Respelling}
\label{subsec:respelling}

@@

The algorithm is simple. In most cases, a phoneme corresponds uniquely to one
character in the respelling. Exceptions are as follows:
\begin{enumerate}
\item{The phone {\em (x)} is spelled {\em ``ch''}.}
\item{The phone {\em (dz) (dzh)} are spelled {\em ``dz'' ``dž''}.}
\item{The diphtongs {\em (aw) (ew) (ow)} are spelled {\em ``au'' ``eu'' ``ou''}.}
\item{
    Sequences {\em (c) (h), (o) (u), (a) (u), (e) (u), (d) (z), (d zh) } are
    spelled {\em ``c'h'', ``o'u'', ``a'u'', ``e'u'', ``d'z'', ``d'ž''}. Note
    though, that the sequence {\em (c) (h)} is purely hypothetical, as it
    contradicts voiced/voiceless assimilation.
}
\item{
    Voiceless alveolar fricative trill is explicated as {\em ``r'{}''} as in {\em
    tř'ít} vs. {\em dřít}.
}
\item{
    Palatal nasal and labiodental nasal are spelled {\em ``n'{}'', ``m'{}''}.
}
\item{Trailing silence is not represented.}
\end{enumerate}

The module includes two-way transduction, although only the one from array of
phonemes to human-readable phonetic respelling is needed in our application.
Still, the user can mark up special-case pronunciation with the apostrophe, like
the sequence of phonemes {\em (c)} and {\em (h)} with the string {\em c'h}. The
need has never occurred during the six years' lifespan of the application.

Note that when encoding into the phonetic respelling, none of
{\em di ti ni dě tě ně} is ever output. The palatal consonants are always
explicitly spelled out and e.g. the sequence {\em (n) (i)} is always spelled
{\em``ny''}

A few examples of words and their phonetic respelling as output by the
algorithm (given the corresponding pronunciation is input as phoneme list):
\begin{itemize}
\item{nic: {\em ňic},}
\item{kdo: {\em gdo},}
\item{disk: {\em dysk},}
\item{dřít: {\em dřít},}
\item{třít: {\em tř'ít},}
\item{auto: {\em auto},}
\item{nauka: {\em na'uka},}
\item{džbán: {\em džbán},}
\item{odželezněný: {\em od'železněný},}
\item{odznak: {\em odznak},}
\item{podzemí: {\em pod'zemí},}
\item{noc: {\em noc},}
\item{tento: {\em tento},}
\item{hangár: {\em han'gár},}
\item{samba: {\em samba},}
\item{tonfa: {\em tom'fa}.}
\end{itemize}

We postulate that the phonetic respelling is natural to all alphabetized native
Czech speakers as a fact without any supporting research, based on experience
alone. The use of apostrophe for distinguishing ambiguities and special cases is
not 100\% intuitive and presents another point where instruction is necessary
for the user to use this feature properly.

\section{Evaluation}

We have presented our web application as a tool that enables gathering precisely
aligned, phoneme-exact transcription from untrained casual visitors. We have
presented measures for reaching this goal but the degree to which it was reached
remains unclear.

We have no gold standard data to measure the quality of our manual
transcriptions. On the contrary, we use the manual transcriptions as gold
standard for the automatic recognition. What we can to, however, is look at some
random samples and try to get a rough idea of how the system performs.

\subsection{Validation by Forced Alignment}

One thing we can examine are the approvals / rejections of the forced alignment.
Of 109640 forced alignment attempts, 3419 have failed, which makes for 3.12\%.
We have manually inspected 20 random failed attempts and came to the following
numbers:
\begin{itemize}
\item{
    11 cases were false negatives, where the transcription was correct and
    should have been accepted,
}
\item{
    4 cases were caused by acoustic irregularities like noise,
}
\item{
    4 cases were true negatives caused by wrongly chosen segment boundaries and
}
\item{
    1 case was true negative caused by wrong transcription.
}
\end{itemize}

Hence, in 25\% of the minimalistic sample, the forced alignment did its job of a
validator and prevented a piece of broken training data from entering the
dataset. In 55\% it was a nuisance and failure, and in the remaining 20\%, it
rejected a valid transcription but prevented a bad training example from
occurring, so we can see this in positive light.

\subsection{Non-Standard Pronunciation}

We can also track how the scenario described in
subsection~\ref{subsec:respelling} is applied. We have looked up four promising
example records in the dynamic dictionary and checked submitted transcriptions
containing them. Table~\ref{tab:pronunc}

\begin{table}[htpb]
\begin{center}
\begin{tabular}{|l r|l r|l r|l r|}
\hline
Correct written form        & \#
	& wrong phonetic pronunc.   & \#
		& correct pronunc.          & \#
			& phon. respelling          & \# \\
\hline
Moody & 2 & {\texttt m o o d i} & 0 & {\texttt m uu d i} & 4 & {\em múdy}, {\em můdy} & 2 \\
Descartes & 2 & {\texttt d e s c a r t e s} & 0 & {\texttt d e k aa r t} & 4 & {\em dekárt} & 2   \\
Weinfurter & 18 & {\texttt v e j n f u r t e r} & 9 & {\texttt v a j n f u r t r} & 16 & {\em vajnfurtr} & 9 \\
Michelangelo & 6 & {\texttt m i x e l a ng g e l o} & 9 & {\texttt m i k e l a n dzh e l o} & 9 & {\em mikelan'dželo} & 4 \\
\hline
\end{tabular}
\caption{Examples of non-standard pronunciation in the manually transcribed data}\label{tab:pronunc}
\end{center}
\end{table}
